---
title: "Integrity practices of global water service providers"
subtitle: "A review of InWASH integrity assessment data"
author: "Claire Grandadam"
affiliation: "Water Integrity Network"
date: "January 2026"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 1
    toc-title: "Contents"
execute:
  warning: false
editor_options:
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

# Introduction

To reach national policy objectives for water and sanitation service
delivery and SDG6, water service providers globally must increase
service coverage and service quality in the face of growing challenges
related to financing, population growth, and climate change. This has
spurred high interest in understanding the drivers of service provider
performance, particularly from service provider managers, policy-makers,
regulators, banks and other funders, as well as sector analysts.

National and global utility benchmarking processes have been used widely
for such analysis, to assess the way service providers operate, and to
orient sector reforms. The most common benchmarking tools tend to focus
on service quality, financial management, and operational efficiency.
Recently, water sector utility benchmarking tools have also been
evolving in part in response to two developments:

-   First, the recognition that performance is affected by governance
    issues and corruption, and the related evidence that anti-corruption
    and integrity management can support utility performance
    (Barreto-Dillon, et al., 2018; WIN, 2021; IDB, 2024). There is a
    need for better tools that can support service providers striving
    for integrity, equity, sustainability, and resilience.

-   Second, growing criticism of the limits and bias in major utility
    benchmarking tools that have primarily focused on efficiency, to the
    detriment of distributional equity (Bhatt, 2024).

Major utility benchmarking tools such as Aquarating and NewIBnet
recently introduced new process indicators focusing on management
practices (in 2022 and 2023 respectively). There is however limited data
available on their use. The Water Integrity Network (WIN) also developed
a set of indicators to assess integrity management practices of service
providers.

This analysis aims to examine anonymised data from applications of the
Water Integrity Network tool since 2018, to asses the tool's relevance
and to understand patterns of use, possible resistance and challenges to
assessing integrity practices, and trends in integrity management
practices.

# Methods

WIN developed integrity indicators for utilities with support from the
Inter-American Development Bank and in partnership with the Consortium
for Water Integrity in Latin America (with SIWI and cewas). This set of
indicators is generally referred to as the Utility Integrity Assessment.
It is part of WIN's InWASH integrity management toolbox for water and
sanitation service providers, an intensive process for prioritising and
managing the integrity risks service providers face in their daily
operations (a process which can also be run without the Utility
Integrity Assessment). The tool is also available separately as an
online survey at <https://www.waterintegritynetwork.net/inwash>.

Early pilots and assessment by WIN indicated that there are patterns in
how good governance and integrity practices are implemented in
organisations in the water and sanitation sectors globally and that
benchmarking can be useful for urban service providers of a certain size
(very small rural service providers tend to not yet have the capacity
and systems for the governance practices described in the tool). This is
the case even though local context and the local regulatory environment
of course influence which governance and integrity practices are
possible and effective for different utilities.

The Utility Integrity Assessment was designed as a short survey
examining 15 indicators categorised under 5 integrity principles:

1.  Tone at the top (with 2 indicators: on leadership and on codes of
    conduct);

2.  Integrity risk assessment (with 1 indicator);

3.  Integrity controls (with 7 indicators, on control of conflicts of
    interest, whistleblowing, recruitment, procurement, disclosure,
    participation, and financial management);

4.  Corrective action (with 2 indicators, on sanctions for staff and on
    sanctions for contractors);

5.  Monitoring (with 3 indicators, on risk monitoring, review of risk
    management functions, and responsiveness to external accountability
    mechanisms).

Service providers use the tool for a self-assessment. They score
themselves for each indicator on a scale of 1 to 5, based on clear
descriptions of the practices that they would have in place for each
score. All indicators are formulated in similar ways. For example, the
second indicator (IRM1.2) examines how effectively a service provider
sets integrity standards and enforces them through a code of conduct or
similar document. The 5 possible scores are:

-   **Score 1:** There is no written code of conduct (or similar
    document) outlining what the Utility expects from staff regarding
    values, rules, standards, and principles.

-   **Score 2:** Between 1 and 3.

-   **Score 3:** The code of conduct (or a similar document) contains
    most of the following elements (some elements may be missing): an
    ethical framework for decision making, generic examples of what
    constitute acceptable and unacceptable behaviour, guidelines on
    reporting problems anonymously, accountability and disciplinary
    policies for unethical behaviour, a list of ethics and compliance
    resources. The code of conduct was not revised in the last 3
    calendar years. There is evidence that the Utility has organised a
    training on the code of conduct in the past, but the training is not
    routinely provided.

-   **Score 4:** Between 3 and 5.

-   **Score 5:** The code of conduct (or a similar document) contains
    ALL of the elements outlined in Level 3. The code of conduct was
    revised at most 3 years ago or more recently, after 1) the most
    recent changes to workplace profiles, including restructuring,
    relocation, changes in key roles or decision-making processes, or 2)
    the most recent changes in the external environment, including
    sector reform, new relevant legislation, changes in government
    strategies or in contractors’ business practices. Code of conduct
    training is routinely given to new employees as part of their
    induction programme.

This analysis looks at available data since 2018, from service providers
that filled the survey online independently, and from service providers
responding offline either as pilot users of the Utility Integrity
Assesment or as part of an InWASH process, and in each of these two
cases receiving support and validation from a WIN or partner expert or
consultant.

The analysis looks at:

-   whether scores differ based on the type of utility, the way the the
    survey was taken, or within utilities (over time or between
    individual survey takers).

-   whether there is significant variation in the practices of service
    providers across regions,

-   whether there are clear trends, strengths, and weaknesses across
    indicators and principles,

```{r  echo = TRUE}

library(tidyverse)
library(gt)
library(gtsummary)
library(knitr)
library(DT)

ia_data <- read_csv(here::here("data/processed/processed_data.csv"))

glimpse(ia_data)

tbl_data_overview <- ia_data |>  
  summarise(
      online_count = sum(on_off == "online", na.rm = TRUE),
      offline_count = sum(on_off == "offline", na.rm = TRUE), 
      completed = sum(last_page > 4),
      incomplete = sum(last_page < 5),
      anon = sum(region == "UNKNOWN"),
      utilities = n_distinct(utility_code,  na.rm = TRUE),
      countries = n_distinct(country_id,  na.rm = TRUE)
  )

```

@tbl-data-overview shows the number of entries assessed, noting how many
were submitted online and offline, how many were complete or not, how
many were anonymous, and the number of utilities and countries
represented.

```{r}
#| label: tbl-data-overview
#| tbl-cap: "Overview of Utility Integrity Assessment data"

tbl_data_overview |>
    gt() |>
    tab_header(title = "Utility integrity assessment",
              subtitle = "Data from 75 surveys") |>
    cols_label(online_count = "Submitted online",
               offline_count = "Submitted offline",
               completed = "Complete surveys",
               incomplete = "Incomplete surveys",
               anon = "Anonymous entries",
               utilities = "Distinct utilities",
               countries = "Distinct countries")
```

# Results

## Survey completeness

```{r  echo = TRUE}

library(ggplot2)
library(ggthemes)
library(ggpattern)

ia_data <- ia_data |> mutate(status = case_when(
    last_page > 4 ~ "complete",
    last_page < 5 ~ "incomplete"
    ))
```

Most respondents were able to complete the survey (over 70%). This could
indicate that the indicators are generally well understood and relevant
for different service providers. There are many possible reasons for not
completing the survey but it possible that there are still categories of
service providers for whom the indicators are not easy to follow,
relevant, or applicable in their context. WIN indicates utility size is
most likely a factor in how relevant the survey is. This could not be
confirmed, nor could a threshold be specified with this data.

@fig-completeness shows that all respondents using the offline tool
(with support or some validation) completed the survey in full and
provided contact information on page 6. Most online respondents also
completed the survey, even without dedicated support, though 12 did
abandon early on and provided information for only up to 3 indicators
out of 15, on page 1 (indicator IRM1.1 and IRM1.2) and page 2 (IRM2.1).

```{r}
#| label: fig-completeness
#| fig-cap: "Integrity assessment completeness by input method (online/offline)"

ggplot(ia_data, aes(
    x = last_page,
    fill = on_off)) +
  geom_bar_pattern(aes(pattern_fill = status),
                   pattern = 'circle') +
    labs(title = "Integrity assessment completeness by input method (online / offline)",
       x = "Last page acted on",
       y = "Count",
       fill = "Online / Offline",
       pattern_fill = "Status") +
  scale_fill_grey(start = 0.4, end = 0.8) +
  theme_minimal()

```

## Main score trends

```{r}
#|eval: true

library(ggplot2)
library(ggthemes)

principle_overview <- ia_data |> 
  summarise(
    mean_allprinciple1 = mean(mean_principle1, na.rm = TRUE),
    mean_allprinciple2 = mean(mean_principle2, na.rm = TRUE),
    mean_allprinciple3 = mean(mean_principle3, na.rm = TRUE),
    mean_allprinciple4 = mean(mean_principle4, na.rm = TRUE),
    mean_allprinciple5 = mean(mean_principle5, na.rm = TRUE)
  )

principle_overview


ia_ordered_regions <- ia_data |> mutate(
  region = factor(region, levels = c("Eastern and Southern Africa", "Middle East and North Africa", "South Asia", "Latin America and Caribbean", "Western Europe", "UNKNOWN" ))
  )


ia_data_long <- ia_ordered_regions |>
    relocate(mean_principle1, .after = last_col()) |> 
    relocate(mean_principle2, .after = last_col()) |> 
    relocate(mean_principle3, .after = last_col()) |> 
    relocate(mean_principle4, .after = last_col()) |> 
    relocate(mean_principle5, .after = last_col()) |> 
      mutate(id = as.character(id)) |>
        pivot_longer(cols = irm11_short:irm53_short,
               names_to = "indicator",
               values_to = "score")


score_overview <- ia_data_long |> 
  mutate(
    indicator_name = case_when(
      indicator == "irm11_short" ~ "Integrity leadership",
      indicator == "irm12_short" ~ "Code of conduct",
      indicator == "irm21_short" ~ "Integrity risk assessment",
      indicator == "irm31_short" ~ "Control of conflicts of interest",
      indicator == "irm32_short" ~ "Whistleblowing",
      indicator == "irm33_short" ~ "Merit-based recruitment and selection",
      indicator == "irm34_short" ~ "Procurement controls",
      indicator == "irm35_short" ~ "Transparency and disclosure",
      indicator == "irm36_short" ~ "Feedback and participation",
      indicator == "irm37_short" ~ "Management and financial controls",
      indicator == "irm41_short" ~ "Sanctions against staff for integrity violations",
      indicator == "irm42_short" ~ "Sanctions against contractors for integrity violations",
      indicator == "irm51_short" ~ "Integrity risk monitoring",
      indicator == "irm52_short" ~ "Reveiw of integrity risk management function",
      indicator == "irm53_short" ~ "Responsiveness to external accountability mechanisms"
    )) |> 
  group_by(indicator, indicator_name) |> 
      summarise(
      mean_scores = mean(score, na.rm = TRUE),
      mode_scores = names(sort(table(score), decreasing = TRUE)[1]),
      sd_scores = sd(score, na.rm = TRUE)
  )

score_overview


ia_data_long_anon <- ia_data_long |> 
   mutate(anonymity = case_when(
  region == "UNKNOWN" ~ "unknown respondent",
  TRUE ~ "known respondent"
))

score_overview_anon <- ia_data_long_anon |> 
  group_by(indicator, anonymity) |> 
  summarise(
    mean_scores = mean(score, na.rm = TRUE),
    mode_scores = names(sort(table(score), decreasing = TRUE)[1]),
    sd_scores = sd(score, na.rm = TRUE)
  )

score_overview_anon


ia_data_long_filteredanon <- ia_data_long |> 
  filter(region != "UNKNOWN")

score_overview_filtered <- ia_data_long_filteredanon |> 
  group_by(indicator) |> 
  summarise(
    mean_scores = mean(score, na.rm = TRUE),
    mode_scores = names(sort(table(score), decreasing = TRUE)[1]),
    sd_scores = sd(score, na.rm = TRUE)
  )

score_overview_filtered 

score_overview_filtered_source <- ia_data_long_filteredanon |> 
  group_by(indicator, on_off) |> 
  summarise(
    mean_scores = mean(score, na.rm = TRUE),
    mode_scores = names(sort(table(score), decreasing = TRUE)[1]),
    sd_scores = sd(score, na.rm = TRUE)
  )

score_overview_filtered_source

score_overview_filtered_regional <- ia_data_long_filteredanon |> 
  group_by(indicator, region) |> 
  summarise(
    mean_scores = mean(score, na.rm = TRUE),
    mode_scores = names(sort(table(score), decreasing = TRUE)[1]),
    sd_scores = sd(score, na.rm = TRUE)
  )


score_variation_filtered_regional <- ia_data_long_filteredanon |> 
  filter(region != "Middle East and North Africa") |> 
    group_by(indicator, region) |> 
      summarise(sd_regions = sd(score, na.rm = TRUE)
  )

score_variation_filtered_regional

ia_region_wide <- score_variation_filtered_regional |> 
  pivot_wider(
    names_from = region,
    values_from = sd_regions,
    names_prefix = "sd_"
  )

ia_region_wide
```

@tbl-principle-overview shows the mean scores aggregated at principle
level. Scores for all principles are close to or above 3. Principle 3
(Integrity controls) is strongest whereas Principle 4 (Disciplinary
action) is weakest, though the differences are relatively small. There
is more variation across indvidual indicators.

@tbl-score-overview shows the mean scores for all the individual
indicators. Service providers scored their management practices highest
for indicator IRM5.3, which assesses responsiveness to external
accountability mechanisms (including external audits and regulatory
requirements). Service providers also scored themselves high for
indicators IRM3.7 (management and financial controls) and IRM3.6
(feedback and participation). This would imply they have strong
budgeting, accounting, and audit processes in place as well as clear
connection policies (IRM3.7). It also implies they have a policy for
public participation, forums for participation, and strong customer
feedback mechanisms in place, which all influence their management
choices (IRM3.6). The areas with weakest scores were IRM3.2 (on
whistleblowing) followed by IRM5.2 (on review of integrity risk
management function and processes) and IRM4.2 (on sanctions on
contractors for non-compliance).

These findings contrast somewhat with analyses of results of other
integrity management processes where service providers have identified
integrity *risks*, in particular risks related to operations and
customer relations. REF Indeed, service providers are rating themselves
relatively highly even in areas where they are still perceving high
risks, most clearly in how they interact with customers. The disconnect
may be linked to regional variations not captured in previous integrity
risk management process analyses, or it could indicate that the
practices described in the highest level of indicator IRM3.6 are not
ambitious enough to mitigate significant risk, or that some utilities
misjudge the scale and effectiveness of their feedback and participation
practices. A weakness of the Utility Integrity Assessment is that it
provides only limited insight on practices to mitigate operational
risks, even though these are perceived as high.

```{r}
#| label: tbl-principle-overview
#| tbl-cap: "Utility Integrity Assessment- Mean scores per principle"

principle_overview |> 
  gt() |> 
  tab_header (title = "Utility Integrity Assessment- Mean scores per principle") |> 
  cols_label(mean_allprinciple1 = "Principle 1 - Tone from the top",
    mean_allprinciple2 = "Principle 2 - Integrity assessment",
    mean_allprinciple3 = "Principle 3 - Integrity controls",
    mean_allprinciple4 = "Principle 4 - Corrective action",
    mean_allprinciple5 = "Principle 5 - Monitoring") |> 
      fmt_number(
      columns = everything(), 
      decimals = 2
      )

```

```{r}
#| label: tbl-score-overview
#| tbl-cap: "Utility Integrity Assessment: Scores per indicator"

score_overview |>
    gt() |>
    tab_header(title = "Utility Integrity Assessment: Scores per indicator") |>
    cols_label(indicator_name = "Indicator",
               mean_scores = "Mean",
               mode_scores = "Mode",
               sd_scores = "Standard Deviation"
               ) |> 
    fmt_number(
      columns = everything(), 
      decimals = 2)
```

## Impact of anonymity

@fig-score-comp-anon shows the mean scores per principle for all
respondents and indicates, with dots, the impact of anonymity on mean
scores. Unknown respondents appear to have assessed themselves more
severely across all indicators except indicators IRM3.1 (on control of
conflict of interest), IRM 3.2 (on whistleblower protection), IRM5.1 (on
integrity risk monitoring), and IRM5.2 (on review of integrity risk
management function and processes)

```{r}
#| label: fig-score-comp-anon
#| fig-cap: "Mean scores across integrity assessment indicators, with comparison of entries by known or anonymous respondents"

ggplot(score_overview, aes(
    x = indicator,
    y = mean_scores)) +
  geom_col() +
  geom_point(data = score_overview_anon, 
            aes(x = indicator, y = mean_scores, group = anonymity, color = anonymity)) +
  labs(title = "Mean scores across integrity assessment indicators, with comparison of entries by known or anonymous respondents",
       x = "Indicator",
       y = "Mean scores", 
       color = "Data anonymity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 80, hjust = 0.8))
```

## Impact of input method, support and survey validation

@fig-score-comp-source shows the differences between scores of known
utilities that responded online and those of known utilities that
responded offline, received some support, briefly justified their
responses, and had their responses reviewed and commented on. There is
marked difference between scores. Offline, verified scores are
significantly lower than online scores, except for indicator IRM5.3 on
responsiveness to external accountability mechanisms. The difference is
most pronounced for indicators IRM5.2 on review of integrity risk
management function and processes, IRM3.1 on control of conflicts of
interest, and IRM3.2 on whistleblowing. The validation process does
therefore have an impact though it is possible that some of the
differences in scores could also be tied to regional differences and
different regulatory requirements on service providers (regional
distribution is not the same across online and offline responses).

```{r}
#| label: fig-score-comp-source
#| fig-cap: "Comparison of mean scores across integrity assessment indicators from known respondents, according to source (online/offline)"

ggplot(score_overview_filtered_source, aes(
    x = indicator,
    y = mean_scores, 
    fill = on_off)) +
  geom_col(position = "dodge") +
  labs(title = "Comparison of mean scores across integrity assessment indicators from known respondents, according to source (online/offline)",
       x = "Indicator",
       y = "Mean scores", 
       fill = "Source (online/offline)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 80, hjust = 0.8)
        )
```

## Regional variations

@fig-score-comp-region shows mean scores of all known utilities and how
these scores differ for utilities grouped by region. It is to be noted
that some regions are not represented as widely as others. Utilities in
Western Europe appear to have somewhat higher scores on many indicators
but other scores are quite mixed. The main differences can possibly be
explained by different policy environments and legal status for service
providers, or different regulatory requirements.

However, as shown in @tbl-score-region, there is still significant
spread in the data for utilities within a region. This suggests other
factors, including possibly different dates of establishment of formal
service provision in a country, size of surveyed utilities, or other
more local context factors could also be influencing the regional
trends. These require further exploration.

```{r}
#| label: fig-score-comp-region
#| fig-cap: "Comparison of mean scores across integrity assessment indicators from known respondents, according to region"

ggplot(score_overview_filtered, aes(
    x = indicator,
    y = mean_scores)) +
  geom_col() +
  geom_point(data = score_overview_filtered_regional, 
            aes(x = indicator, y = mean_scores, group = region, color = region)) +
  geom_line(data = score_overview_filtered_regional, 
            aes(x = indicator, y = mean_scores, group = region, color = region)) + 
  labs(title = "Comparison of mean scores across integrity assessment indicators from known respondents, according to region",
       x = "Indicator",
       y = "Mean scores", 
       color = "Region") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 80, hjust = 0.8))

```

```{r}
#| label: tbl-score-region
#| tbl-cap: "Utility Integrity Assessment: Variation in regions"

ia_region_wide |> 
    gt() |>
    tab_header(title = "Utility Integrity Assessment: Variation in regions") |>
    fmt_number(
      columns = everything(), 
      decimals = 2)
```

# Conclusions

The data from Integrity Assessments is still too limited to confirm
decisive trends in integrity management practices of water and
sanitation service providers. However, the data does still suggest that
integrity management practices could be improved, especially to mitigate
conflicts of interest, take action against non-compliant contractors,
and protect whistleblowers.

Many organisational and sector anti-corruption processes focus on
procurement control. This is an important area of risk though it is not
the weakest management area and some progress appears to already have
been made.

The completeness of responses, the spread of scores, and the spread of
scores even within regions suggest the tool is able to capture practices
that are influenced by the management of individual utilities, and not
just broad contextual trends. The tool can thus provide important
insight for utilities (and for regulatory authorities) on paths for
action to control major integrity risks.

# References
